---
title: "Comparing network clusters from tragula with expert-curated clusters"
author: "Art Poon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir="~/git/tragula")
```

Load information about department faculty, where I have manually added annotations on the membership of faculty members in 12 areas of research.  These areas were identified through questionnaires and manual assessment of clusters.
```{r}
palm <- read.csv("data/palm.tsv", sep='\t')
for (i in 5:ncol(palm)) {
  palm[is.na(palm[,i]), i] <- FALSE
}
```

Next, we'll load the Wasserstein distance matrix that was pre-computed for the main README document.
```{r warning=FALSE, message=FALSE}
load("results/palm_wdist.RData")

# make sure labels match the distance matrix
palm$filename <- paste(palm$lastname, palm$forename, sep="_")
palm$filename <- gsub(" ", "_", palm$filename)
idx <- match(attr(wdist, "Labels"), palm$filename)
labels <- palm$initials[idx]
```

Let's start with the original network community detection analysis of these data:
```{r fig.width=5, fig.height=5, dpi=150, out.width="66%"}
source("scripts/visualize.R")
g <- make.knn(wdist, k=3)

require(igraph, quietly=TRUE)
cl <- cluster_louvain(g)
par(mar=rep(0, 4))
plot(g, mark.groups=cl, vertex.shape="none", vertex.label=labels, 
     vertex.label.cex=0.7, vertex.size=6, 
     edge.width=2, edge.arrow.mode='-')
```

A significant issue is that there are an enormous number of ways to cluster these data, not only by varying the criteria for drawing edges (distance threshold and maximum in-degree), but also by changing the community detection method applied to the graph:
```{r fig.width=5, fig.height=5, dpi=150, out.width="66%"}
res <- cluster_label_prop(g)
sum(sapply(groups(res), length))
par(mar=rep(0, 4))
plot(g, mark.groups=res, vertex.shape="none", vertex.label=labels,
     vertex.label.cex=0.9, vertex.size=6,
     edge.width=2, edge.arrow.mode='-')
```

## Measuring concordance
Let's calculate the adjusted Rand index for concordance of partitions with potentially overlapping communities (omega index, Collins and Dent 1988):
$$\omega(C_1, C_2) = \frac{\omega_u(C_1, C_2) - \omega_e(C_1, C_2)}{1-\omega_e(C_1, C_2)}$$
where
$$\omega_u(C_1, C_2) = \frac{1}{M} \sum_{j=0}^{\max(K_1, K_2)} | t_j(C_1) \cap t_j(C_2)|$$
is the unadjusted omega index with $M={n\choose 2}$ node pairs, $K_1$ is the number of communities in partition $C_1$, and $t_j(C)$ is set of node pairs that appear exactly $j$ times in partition $C$ (Xie, Kelley and Szymanski, https://arxiv.org/abs/1110.5813v4).
The expected value is calculated as:
$$\omega_e(C_1, C_2) = \frac{1}{M^2} \sum_{j=0}^{\max(K_1, K_2)} |t_j(C_1)| \cdot |t_j(C_2)|$$
