---
title: "Generating networks of shared research expertise"
author: "Art Poon"
date: 'Departments of Pathology and Laboratory Medicine; Microbiology and Immunology; Computer Science'
output: 
  revealjs::revealjs_presentation:
    self_contained: true
    css: Rmd-style.css
    center: true
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Expertise mapping

* A common task for academic units is to identify clusters of research interests/expertise
  * Recruiting prospective graduate students
  * Facilitating interactions among faculty
* This task is often done manually, which is subjective, tedious, and time-consuming.

---

## Objectives

* Develop an automated workflow to:
  * extract research themes from each member's contributions to the peer-reviewed literature
  * use multivariate analysis to determine each member's relation to their  peers based on shared research themes
  * develop visualizations of analysis outputs
* Release as an open source project that might be used by other groups

---

## Database

* NCBI PubMed can be accessed through the Entrez API
  * Results are returned in an XML format by default
* Biopython provides an interface to this API and XML parsing functions

```python
>>> from Bio import Entrez
>>> Entrez.email = "apoon42@uwo.ca"
>>> response = Entrez.esearch(db="pubmed", term="Art Poon")
>>> result = Entrez.read(response)
>>> result
{'Count': '112', 'RetMax': '20', 'RetStart': '0', 'IdList': ['38485563', '38355869', '37547379', '37463439', '37292785', '37187604', '37101658', '36846827', '36809686', '36632480', '36449514', '36395093', '36178092', '35916373', '37552744', '35500121', '35202429', '34941890', '37124703', '34578305'], 'TranslationSet': [{'From': 'Art Poon', 'To': 'Poon, Art[Full Author Name]'}], 'QueryTranslation': 'poon, art[Author]'}
```

---

## Data collection

* Retrieve title, keywords and abstract for all PubMed records associated with author query
  * Exclude titles starting with "Author Correction" or "Correction"
* Manually compose query to exclude other authors
  * *e.g.*, `Cameron, Lisa NOT Cameron, LA NOT Cameron LJ`
* Write results to a JSON file for each member.

---

## Natural language processing

* Using Python module `nltk` (natural language toolkit)
* Word tokenizer breaks strings down into words and punctuation.
* Part-of-speech tagging categorizes words by their meaning and context.
* Exclude punctuation, coordinating conjunctions and other words in a manually-curated list, *e.g.*, "when"

---

## Term-document matrix

* Count the co-occurrences of words per document (title+keywords+abstract)

```{r echo=FALSE}

```


